Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-07-02 MIMOSA.tex
update si
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\begin{equation}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
\begin{equation}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:beta_prior}
\end{equation}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal. Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder.

@@

\subsubsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
We fit the multivariate MIMOSA model to the two--sided Fluidigm data, looking at coexpression of pairs of genes in single cells across the four stimulations. In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B). Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation. The typical approach to analyzing polyfunctional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases. There is no difference in the marginal counts for any sample (Figure~\ref{fig:polyfunctionality} C). In contrast, multivariate MIMOSA tests all cell subpopulations simultaneously, and identifies a significant difference between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the individual combinations, which requires multiplicity adjustment and a potential loss of power.

@@

>>>>

\begin{equation}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
\begin{equation}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:beta_prior}
\end{equation}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal. Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder. In the supplementary material we describe modifications for an alternative constrained model where $p_s>p_u$ (supplementary material~\ref{supp:constrained}).

@@

\subsubsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
We fit the multivariate MIMOSA model to the two--sided Fluidigm data, looking at co--expression of pairs of genes in single cells across the four stimulations. In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B). Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation. The typical approach to analyzing polyfunctional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases. There is no difference in the marginal counts for any sample (Figure~\ref{fig:polyfunctionality} C). In contrast, multivariate MIMOSA tests all cell subpopulations simultaneously, and identifies a significant difference between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the individual combinations, which requires multiplicity adjustment and a potential loss of power.

@@

\subsection{Constrained Beta--binomial model}
\label{supp:constrained}
We can define a model where we constrain the stimulated proportions under the alternative model such that $p_s>p_u$. In this case, the only changes required are for the alternative marginal likelihood $\mathrm{L}_1$ defined in equation~\eqref{model2:unconstrained}. The marginal likelihood for the constrained model becomes $\mathrm{L}_1\times \mathrm{C}$, where $\mathrm{C}$ is a ratio of normalizing constants for the likelihood and the prior.

\begin{align}
\begin{split}
	 \mathrm{C}=\frac{\int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}p_{ui}^{n_{ui}+\alpha_u-1}(1-p_{ui})^{N_{ui}-n_{ui}+\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(N^i_s-n^i_s+\beta_s,n^i_s+\alpha_s)\right)d p_{ui}}{   \int\limits_{p_{ui}=0}\limits^{1}\left(\frac{1}{\mathrm{B}(\alpha_u,\beta_u)}p_{ui}^{\alpha_u-1}(1-p_{ui})^{\beta_u-1} \right)\left(\mathrm{I_{1-p_{ui}}}(\beta_s,\alpha_s)\right)d p_{ui}}
\label{eq:model2post}
\end{split}
\end{align}

The term $I_{1-p_{ui}}(\beta_s,\alpha_s)=1-I_{p_{ui}}(\alpha_s,\beta_s)=Pr(p_{si} > p_{ui}; \alpha_s,\beta_s)$ is the regularized incomplete Beta function, which is just the CDF of a Beta distribution with parameters $\alpha_s,\beta_s$, resulting in a 1--dimensional integral that can be computed via Monte--Carlo integration (i.e. in our EM implementation). However, this can be computationally costly in the MCMC framework. In the latter case, we can compute the integral exactly if $\alpha_s$ and $\beta_s$ are integers by using the identity:
\begin{align}
I_{1-p_u}(\beta_s,\alpha_s) = \sum_{j=\beta_s}^{\beta_s+\alpha_s-1} \frac{(\beta_s+\alpha_s-1)!}{j!(\beta_s+\alpha_s-j)!}(1-p_u)^j(p_u)^{\beta_s+\alpha_s-j}
\label{eq:IBident}
\end{align}
Typically, in ICS data $\beta_s>>\alpha_s$, leading to relatively few terms in the sum in equation~\ref{eq:IBident}.  This leads to an exact expression for the normalizing constant:
$$
foo
$$
